---
title: "Monday Assignment"
output: html_document
author: "Anneli Tandberg"
date: "2022-09-28"
---

# Preparation, loading and getting help

### Working Directory

Using the setwd function will set your working directory to the specified folder. The working directory allows R to access data in that folder directly, without needing to specify a path.

```{r}
#setwd("~/Desktop/R-kurs")
```

If you don't set your working directory, you will have to write the complete path for R to find the object you're trying to load
```{r}
#a <- ("~/Desktop/R-kurs/datafile.RDS")
```

Instead of just writing the file name that is in that working directory
```{r}
#b <- "datafile.RDS"
```

### Packages

Packages are pre-made functions to be used. One can download a package for R by simply using the install.packages function. You have to load the package for each R script with the library function
```{r}
#install.packages(tidyverse)
library(tidyverse)
```


### Help function

The help function sends you directly to the help page for any function in R

```{r}
help(help)
help("install.packages")
```

The help functions are not always very useful, so you should also use Stackoverflow or google

### Assigning values

The assign function allows you assign something to an object. An easier way of doing this is to use the arrow symbol

```{r}
assign("c", 3)
c <- 3
```

### Viewing files

To view a file, use the view function. 

```{r}
view(iris)
```

Iris is an already existing dataset that comes with R

### Loading and removing objects

In order to load your data, you can use the read function. What comes after depends on the file type. Here we read a CSV-file
```{r}
#read.csv("~/Desktop/datafiles/fil1.csv")

```

If you wish to remove something from your environment, you can use the rm function
```{r}
rm(c)
```

# Object types

There are several types of objects in R. One such object is a vector. Vectors are sequences of units (such as numbers for height) that are of the same type (no mixing height and name). Use the c function to combine the elements

```{r}
vector_example <- c(9, 11, 4, 22, 85)
```

To get a sequence of numbers or letters for a vector, use the : to specify the range
```{r}
vector_example <- 1:4
```


One can check logical statements, which are either TRUE or FALSE
```{r}
vector_example > 3
vector_example < 3
```


Objects can be of, for example, type interger (whole numbers) or "strings", which are letters or words. R recognizes that something is a string and not a function by using " "
```{r}
d <- c("male", "female", "male", "male")
```

To check the type of an object, use the type of function
```{r}
typeof(vector_example)
```

To check the length of an object, use the length function
```{r}
length(vector_example)
```

To get hold of a single value in a data set, use the square brackets []
```{r}
d[3]
```


# Creating dataframes

When you have the variables you need, it's useful to combine them to one table consisting of rows and columns. First we make some fake data
```{r}
breed <- c("german shepherd", "pomerian", "dalmantinian", "poodle", "golden retriever", "border collie") 
age <- c(2, 3, 11, 9, 2, 6)
lovely <- c("yes", "yes", "yes", "yes", "yes", "yes")
sex <- c(0, 0, 1, 1, 0, 0)
```

Now we combine these lonely variables to a single data frame with the data.frame function
```{r}
dogs <- data.frame(breed, age, lovely, sex)
```

Let's get an overview of this beauty with the summary function
```{r}
summary(dogs)
```
I want the mean age for only the male (0) dogs. Use the mean function and [] to grab hold of something specific, and $ to tell R that it's a variable in a specific dataset. Double equal sign means that the object must be exactly equal to something, in this case 0, which is the code for male
```{r}
 mean(dogs[dogs$sex=="0","age"])
```

I want to standard deviation for the age of female dogs. I use the sd function
```{r}
sd(dogs[dogs$sex=="1","age"])
```

# Creating visuals

### Tables

Creating tables in R is difficult and tricky. There is an online tool that can help: [helpful link](https://www.tablesgenerator.com/markdown_tables)

|   | A  | B | C |   |
|---|----|---|---|---|
| D | 1  | 2 | 3 |   |
| E | 2  | 0 | 4 |   |
| F | 11 | 4 | 1 |   |

Why don't the letters align? I don't know.

The pander package is better for making tables in R. I'm going to use the cars data set which is already in R, but it's long and will give us a terrible table. Use the head or tail function to specify that you only want the start (n = how many) or the end of the data set as a table

```{r}
library(pander)
data(cars)
pander(head(cars, n=5))
pander(tail(cars, n=5))
```

### Histogram

To create a histogram, use the hist function
```{r}
hist(dogs$age)
```




# R-markdown

R-markdown lets you combine free text and R-functions to a neat html file. Use # for a big header, and the more # you use, the smaller the header

# Header 1

## Header 2

### Header 3

Use the [] followed by () to create a [fancy link](https://www.nasa.gov/)

Use the ALT + CMD + I to make a field where you can use R normally, which will also show in the final product
```{r}
this_is_a_R_field <- TRUE
```

# R as a calculator

Not the most shocking, but R can be used as a calculator
```{r}
e <- 4
f <- 10
e+f
e*f
e/(f*10)
```

# Logical operators

< less than
<= less than or equal to
 > more than 
 >= more than or equal to
!= not equal to

