---
title: "day 2"
output: html_document
date: '2022-09-27'
---

# DAY 2

## Loops - a way of repeating actions

Starts with the general for(b in 1:5){ print(b) }

```{r}
b <- c(1,2,3,4,5,6,7,8,9,10)
string <- c("hei", "du", "se")
for (b in string) {
  print("hei")
}

s <- c("some", "character", "for", "the", "example")
for (b in 1:5) {
  print(s[b])
}
```

```{r}
for (b in 1:5) {
  print(b)
}
```

## Working with data

### Loading data in R (not creating, as done yesterday, but importing e.g. spss data)

csv files r files

RDS format - binary file, open format, the official R format (in addition to R data files). Can save only one object.

Rdata - multiple objects

```{r}
help("readRDS")
saveRDS(iris, "testdata.RDS")
```



```{r}
rm(list = ls())
ls() #list obkects in the environment
rm() #clearing the environment

```

Importing csv files

read.csv = comma separated

read.csv2 = semicolon separated

unlink can be used to remove a file in the WD

assign(x = df2, value = readRDS("test.RDS")

```{r}
data("airquality")
View(airquality)
dim(airquality)
str(airquality)

airquality2 <- subset(airquality, Day==1, select = -c(Temp, Solar.R)) #subsets all columns EXCEPT for the two specified and selects observations in day 1 only
airquality3 <- subset(airquality, select = c(Temp, Solar.R)) #subsets two columns

colnames(airquality) #gives the colnames of the dataset
help(colnames)
```

```{r}
library(readstata13) #recommended for stata files
library(haven) #recommended for spss files
library(foreign) #recommended for spss files
```

## Data classes
A factor = a categorical variable - can be string or numerical (?)

```{r}
class(airquality$Temp)
str(airquality$Temp)
is.factor(airquality$Temp)
```

## Tidyverse/Excel

Excel files can be complicated since it may have several sheets. Can specify the following way:
df <- readxl::read_excel("path/to/file.xlsx",

sheet = 1)

Newer excel format = .xlsx

```{r}
library(tidyverse)
library(devtools)
library(readxl)
```

```{r}
haven::write_sav(data=airquality, path = "airquality.sav")
haven::read_sav(file="airquality.sav")

```

## Summary
```{r}
summary(cars)
summary(cars$speed)
plot(cars$speed, cars$dist)
hist(cars$speed)
```


```{r}
data("attenu")
if(is.numeric(iris$Species)) {
  mean(iris$Species)
}else{
  str(iris$Species)
}
sum(is.na(attenu$station))
```

length will not work since its a logical (i.e. 0 or 1) - meaning that sum must be used.


## Character vectors

Learn important printing functions, e.g.

-   Paste(), cat(), paste0()

-   substrings of a character vector

    -   each character in a string has a number, which can be checked with nchar(vector)

-   stringsplit

-   which can be used for example if you are looking for NAs and you get true/FALSE - you can ask WHICH is na

```{r}
help(paste)
state.name
substring(state.name, 1, 3)
?substring
?which

```

## Stringr 
Part of tidyverse - can be used for strings/character vectors. 
7 important verbs:
1. str_detect()
2. str_count()
3. str_subset()
4. str_locate()
5. str_extract()
6. str_match()
7. str_replace()

tidyverse always use _ in the function names

## Regular expressions
Best to learn using the regexplain package



