---
title: "day1 - Heidi Umbach Hansen"
output: html_document
date: '2022-09-26'
---

```{r, warning=FALSE}
library(rmarkdown)
library(knitr)
```

# The basics

## Accessing help pages

R have very handy help pages, which can be accessed in the following ways:

```{r}
?help
help("mean")
?mean
?citation
```

## Assigning values

```{r}
A <- 2
B <- 5
x <- c(1:10)
```

# R markdown

Various ways to creating headings, e.g. with \# (number of \# creates subheadings), = or -. NB! When using = the next line must be empty.

# R markdown

## Inline formatting

*italic* *italic*   
**bold** __bold__   
sub~s~cript super^s^cript  
inline `code`  
 

## Adding chunks

Warning=FALSE will not display the messages etc

```{r, warning=FALSE, message=FALSE}
library(tidyverse)  #loading the tidyverse package
vector_test <- c(10,20,30,40,50)  #Creating a vector
assign(x="A", value = 10)
print(A)
```

## Making comments
The following will not be rendered
<!--
An interesting comment
-->

## Diverse

I Rmarkdown så er environment tomt - det er et slags lokalt environment som blir generert når man trykker "Knit". Det som gjør at det blir reproducable - siden alle objekter må lages i RMarkdown fra scratch.

Er ikke mulig å bytte mellom lokalt og globalt miljø

## Tables 
Can be generated online: <https://www.tablesgenerator.com/markdown_tables>

Or use pander

```{r}
library(pander)
```

# Working with dynamic r tables

Using the package 'pander' is a smarter way to work with dynamic tables

```{r}
data(cars)
pander(cars)
pander(head(cars, n=3L))
```

NB! Ved å skrive eks. 6L istedenfor 6 (begge er integer), så sier man at det ikke er noen desimaler, og er mer minnebesparende (og kan dermed gå fortere).

# Task
```{r}
data(cars)
hist(cars$speed)
help(hist)
```


