---
title: "PSY9510-Day2"
author: "Heemin Kang"
date: '2022-09-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## For-loop

```{r}
s <- c("a", "b", "c", "d", "e")
for (b in 1:5){
  print(s[b])
}
```

# Work with data

## RDS _vs_ RData
  - RDS is used for saving (`saveRDS(object, file)`) and loading (`readRDS(file)`) a single object

```{r}
data(iris)
saveRDS(object = iris, file = "testdata.RDS")   # File is saved in the current working directory
a <- readRDS("testdata.RDS")                    # Assigning RDS file to a variable
a
#assign(a, readRDS("testdata.RDS"))             # Alternative way for assignment
#dim(a)
#str(a)
```

  - RData is more general and can include more than a single object

## Cleaning workspace
```{r}
rm(list=ls())                                   # ls returns a vector of character strings giving the names of the objects in the specified environment
```

## CSV data format
- CSV: Comma Separated Values
  - Semicolon or tab could also be delimiter but requires a different syntax
  - `read.csv`: Reads a file (even directly from a URL) in table format and creates a data frame from it
    - args:
      - file (character): file name
      - header (logical): whether the file contains the names of the variables of its first line
      - sep (character): delimiter ("," by default)

```{r}
write.csv(iris, file = "irisTest.csv")
unlink("irisTest.csv")                          # deletes the file(s) or directories specified by x
```

## Fixed-width input files
```{r}
#help(write.fwf)
```

## Subsetting using `subset`
- subset (logical): elements or rows to keep: missing values are taken as false.
- select: columns to select from a data frame.
- drop: passed on to [ indexing operator
```{r}
data(airquality)
airquality$Temp > 80                                  # returns logicals if a specified column meets the given condition

subset(airquality, Day == 1, select = -c(Temp, Wind))

colnames(airquality)
rownames(airquality)
```

## Reading Stata data

library(readstata13)
library(foreign)
library(haven)

df <- readstata13::read.dta13("path/to/file.dta")  # '::' specifies a function in a specific package
df <- foreign::read.dta("path/to/file.dta")
df <- haven::read_dta("path/to/file.dta")

# Factors
The function factor is used to encode a vector as a factor (the terms ‘category’ and ‘enumerated type’ are also used for factors). If argument ordered is TRUE, the factor levels are assumed to be ordered. For compatibility with S there is also a function ordered.

```{r}
is.factor(airquality$Temp)                                        # returns logical value if the given column is a factor
```


## Reading excel file using `readxl`
The readxl package supports .xls and .xlsx

```{r}
library(readxl)
library(tidyverse)

readxl_example()                                                  # displays the list of examples
xlsx_example <- readxl_example("datasets.xlsx")   
read_excel(xlsx_example)                                          # reads both .xls and .xlsx files
excel_sheets(xlsx_example)                                        # lists the sheet names
read_excel(xlsx_example, sheet = "chickwts", range = "B1:D2")     # a worksheet can be specified by number as well

```

# Summary function
```{r}
summary(attenu)
vec <- colnames(attenu)
sum(is.na(attenu$station))                                                # checking the number of missing values

if (is.numeric(attenu$event)){
  mean(attenu$event)
} else {
  str(attenu$event)
}
```

```{r}
data(iris)
if (is.numeric(iris$Species)){
  mean(iris$Species)
} else {
  str(iris$Species)
}
```

# Plot
## Scatterplot
For creating a scatter plot using `plot`
```{r}
plot(cars$speed, cars$dist)

vec2 <- cars$speed
is.character(vec)
for (pasta in vec2){
  print(pasta)
}
```

Substring
```{r}
fn <- "tofu"
nchar(fn)
substr("cooked tofu", start=1, stop=2)
substr("cooked tofu", start=8, stop=12)
strsplit("cooked tofu", split = " ")
```

```{r}
which(LETTERS == "R")
which(ll <- c(TRUE, FALSE, TRUE, NA, FALSE, FALSE, TRUE))
names(ll) <- letters[seq(ll)]
which(ll)
```

## Some more stringr-functions
-str_detect()
-str_count()
-str_subset()
-str_locate()
-str_extract()
-str_match()
-str_replace()

# How to do nchar(), concatenate, and substring() with **stringr**
```{r}
library(stringr)
x <- c("why", "video", "cross", "extra", "deal", "authority")
str_length(x)
str_c(x, collapse = ", ")
str_sub(x, 1, 2)
```

# Regular expressions
[RegExplain](https://github.com/gadenbuie/regexplain)
```{r}
str_subset(x, "[aeiou]")
library(regexplain)
regexplain_gadget()
```



