---
title: "PSY9510-Day1"
author: "Heemin Kang"
date: '2022-09-26'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# *Introduction to the R language*

## Set-up

- `source(filename)` causes _R_ to accept its input from the named __file__ or __URL__. Input is read until the end of the file, then the lines are evaluated sequentially in the chosen environment.
- `setwd(path)` sets the passed argument as a working directory
- Two types of _Environment_: Global (accessible in the current working space) vs. local
- *Comment*  along the codes in consistent structure for better readability!
- `help(function_name)` or `?(function_name)` opens <help> document
- `str(obj)` compactly displays the internal structure of an R object. More general than typeof()

## Installing and loading packages

`install.packages("tidyverse")`
```{r, message = FALSE}
library(tidyverse)             # load the tidyverse library
```

## Assigning value(s) to a variable
```{r, message = FALSE}
assign(x = "variable", value = 5)

I <- c(1, 2, 3)                # c(...): concatenate the passed arguments
S <- c("A", "B", "C")
B <- c(1, 2, "A")              # coercing data type numeric to character

TF <- c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE)
str(TF)
length(TF)

rm(variable)                   # removes a given variable for environment

```

 
## Creating a list
- List 1
- List 2
  - List 3
  - List 4

## Creating a table
[tablesgenerator.com](https://www.tablesgenerator.com/markdown_tables)

| A | B |
|---|---|
| 1 | 2 |
| 3 | 4 |

## Working with dynamic R tables

As you see, Rmarkdown's tables are stupid and we want a smarter way for working with dynamic tables. For this, we use the `pander` R package.

```{r}
library(pander)
data(cars)
#pander(cars[c(1:5),])
pander(head(cars, n = 5L))
pander(tail(cars, n = 3L))     # L just means integer    
```

A = 6    # 6.0000000...  
B = 6L   # 6                 

```{r}
data(iris)
pander(iris, n = 4)
```

## Creating a dataset

We can create a data frame using the `data.frame` function

```{r}
#help("data.frame")
```

Let's print all the English letters. It's defined in R automatically and we do not need to define it ourselves
```{r}
LETTERS
L3 <- LETTERS[1:3]             # [] subsets a vector by the given indices
```

## Subsetting data in R

```{r}
data(iris)
print(iris[10:15,])
```
Select the rows 12:14 and columns 2:5
```{r}
pander(iris[12:14, 2:5])  # pander() generates a more readable table than print()
```

You can also be very selective in indexing
```{r}
#pander(iris[c(1, 3 ,5 , 6, 1), 2:5])
dim(iris[iris[1]>5,])[1]
```


```{r}
name = c("Emily", "Julie", "Stina", "Eirik", "Nadine", "Sara", "Ole", "Anders", "Niklas", "Fredrik", "Haghish")
age = c(27, 25, 31, 26, 31, 22, 27, 37, 44, 45, NA)
gender = c("female", "female", "female", "male", "female", "female", "male", "male", "male", "male", "male")


df <- data.frame(name, age, gender)
```

**QUESTION**: What is th mean of age for female participants of the class PSY9510?

```{r}
df[df$gender == "female", "age", drop = FALSE]
mean(df[df$gender == "female", "age"])
```
Let's assign the mean to an R object
```{r}
femmean <- mean(df[df$gender == "female", "age"])

```
The analysis showed that the mean age of females in this data set is `r femmean`. **This way, we can bring R objects inside text**.


Let's calculate the standard deviation for male's age:

```{r}
# help(sd)
sd(df[df$gender == "male", "age"], na.rm = TRUE)
```


How can I evaluate if the age of the participants is above or equal to 35 or not?

```{r}
val <- df$age >= 35

df$age35 <- val
pander(df)
```


We can add figures to R Markdown. Here I will add a histogram using the `hist` function.

```{r}
data(cars)
hist(cars$speed, size = c(5,5))
```

