---
title: "Day 2"
author: "J.M. Lengle"
date: "2022-09-28"
output: html_document
---

# **Introduction to statistics in R: 2**

# *Follow-up of day 1*

## Loops

* Many uses
   + we can use loops to subset
   + or to repeat an action for a number of variables
      + f.ex.taking mean of every variable in a set, changing column names

```{r}
for(b in 1:5){print(b)}
s <- c("some", "character", "for", "the", "example")
for(i in s){print(i)}
for(indiancurry in s){print(indiancurry)} #we can use any word as the i in a for() loop
for(b in 1:5){print(s[b])}
s[3]
```

* We can also use loops for *if-else statements*

```{r}
A <- 6
if(A<2){print("small")} else {print("big")}
ifelse(s!="example", " ", s)      #use ifelse() for vectors
```

# Working with Data
  + Loading data into R
  + Writing data in different formats
  + Summarizing data

## Loading data into R

* R has a number of built-in datasets that can be accessed with functions, f.ex. `data(cars)` or `data(iris)`
* R can also read a dataset stored in computer with `readRDS()` or `saveRDS()`
  + To use `readRDS()` you only need the file
  + To use `saveRDS()` you need the object and the file: (name of R file to save, desired file name)
      + In RDS you can save only one file
      + RDS is a binary data type. "object to serialize"=to make binary
      + RData is more general and can include many objects in a single file

```{r eval=FALSE, echo=TRUE}
help(readRDS)
data(iris)
saveRDS(iris, "testdata.RDS")   #save a file in your current WD
assign(x="df2", value = readRDS("testdata.RDS"))
df2 <- readRDS("testdata.RDS")
```

+ You can view a list of objects in the workspace or clear the R workspace within the RStudio GUI, or by typing `ls()` or `rm(ls())`, respectively
   + Typing `ls()` with no object, causes R to list all objects you have
       +`character(0)` means nada

```{r eval=FALSE, echo=TRUE} 
ls()                            #list objects 
rm(ls())                        #remove objects from the workspace
```

### CSV data format
  + Comma separated
  + Most popular type of data, can be read in a text file
  + Every row of the data ends with a new line
  + CSV file can also be a complete URL
  
```{r eval=FALSE, echo=TRUE} 
help(read.csv)  
data(iris)
write.csv(iris, file="iristest.csv")  #create csv file
help(unlink)
unlink("iristest.csv")   #delete file from wd
unlink("testdata.RDS")
```

## Using data from the internet

+ R can access data from the internet and read the data without storing it on the disk
+ R can read files directly from html links

```{r eval=FALSE, echo=TRUE}
help(download.file)       #use download.file to download from the internet
```

## Reading STATA data

df <- readstata13::read.dta13("path/to/file.dta") # :: use this package specifically

```{r eval=FALSE, echo=TRUE} 
install.packages("readstata13")
install.packages("foreign")
```

```{r eval=FALSE, echo=TRUE} 
??read.stata
library(readstata13)
help("readstata13")
```

### Exploring data

```{r eval=FALSE, echo=TRUE} 
data(airquality)
View(airquality)              #view the data
dim(airquality)               #inquire about data dimensions
str(airquality)               #inquire about structure of data
help("subset")
airquality$Temp>80            #extract observations where temp>80
subset(airquality, Temp>80, select=c(Ozone,Temp)) #show ozone and temp for data points where Temp>80
subset(airquality, Day==1, select=-Temp) #drop/exclude Temp from data set on day 1
subset(airquality, select=Ozone:Wind) #show columns ozone through wind
subset(airquality, select= -c(Temp,Wind)) #drop/exclude both Temp and Wind from data set
colnames(airquality) #retrieve column names
rownames(airquality) #retrieve row names
help("colnames") 
```

+ If loading STATA or SPSS data into R, be extra careful with factors

```{r eval=FALSE, echo=TRUE} 
help(factor)
is.factor(airquality$Temp)  #is airquality$Temp a factor? T/F
class(airquality$Temp)      #what is the class?
str(airquality$Temp)        #what is the structure?
```

## Reading Excel sheets

+ Tidyverse: collection of R packages including readxl

```{r eval=FALSE, echo=TRUE} 
library(readxl)
help("read_excel")
readxl_example()
readxl_example()
readxl_example("clippy.xls")  #example file from readxl package
xlsx_example <- readxl_example("datasets.xlsx")
read_excel(xlsx_example)      #reads both xls and xlsx files
xls_example <- readxl_example("datasets.xls")
read_excel(xls_example)       #use `print(n = ...)` to see more rows
excel_sheets(xlsx_example)    #list the excel sheet names
read_excel(xlsx_example, sheet = "chickwts") #specify a sheet name by name or number
read_excel(xls_example, sheet = 4)
read_excel(xlsx_example, n_max = 3) #specify the sheet and cell range
read_excel(xlsx_example, range = "C1:E4")
read_excel(xlsx_example, range = cell_rows(1:4))
read_excel(xlsx_example, range = cell_cols("B:D"))
read_excel(xlsx_example, range = "mtcars!B1:D5")
read_excel(xlsx_example, na = "setosa") #if missing cells are represented by something else than NA, set the NA argument
```

### Write an SPSS file from an R object

```{r eval=FALSE, echo=TRUE} 
data("airquality")
haven::write_sav(data=airquality, path="airquality.sav")
```

## Read an SPSS file into R

```{r eval=FALSE, echo=TRUE} 
??read.spss
df <- haven::read_sav(file = "airquality.sav")
str(df)
library(pander)
pander(head(df))
```

```{r eval=FALSE, echo=TRUE}  
help(summary)
data(attenu)
View(attenu)
colnames(attenu)
vec <- colnames(attenu)   #create a vector that consists of the column names
class(vec)              
vec[3]                    #extract the third column name
vec2 <- vec[3:1]          #extract column names 3,2,1 in that order
summary(attenu, digits=4) #summary data on data columns
is.na(attenu)             #T/F about whether data=missing for each observation
help(is.na)
```

```{r}
help("is.numeric")
if (is.numeric(attenu$event)==TRUE){mean(attenu$event)} #calculate mean of all numeric values in column events
if (is.numeric(attenu$event)){mean(attenu$event)} else {str(attenu$event)} #calculate mean of all numeric values in column events, otherwise describe data structure
help("mean")
if(is.numeric(iris$species)){mean(iris$species)} else {str(iris$species)} 
is.na(attenu$station)         #list of T/F statements
length(is.na(attenu$station)) #this doesn't work since it counts all statements in the vector, i.e. both true and false
sum(is.na(attenu$station)) #sum of true and false gives number of missing observations since true 1 and false 0
help(sum)
```

# Create a scatter plot

```{r}
plot(cars$dist) #create scatterplot. index and number. index is the number of the observation
cars$dist
plot(cars$speed,cars$dist) #specified the x and the y
```

# Working with character variables

+ Important printing functions
  + `paste()`
  + `cat()` (can't assign to default)
  + `print()`
*Note: atomic data structures can only have one type of data. Vectors are atomic data structures and can only have one data type. Thus, when you evaluate whether a vector is character or number or not, you only get one T/F, not an evaluation for every element of the vector.*

```{r eval=FALSE, echo=TRUE} 
help("paste") #paste function automatically concerts vectors to character
help(as.character) #converts to character
help ("cat")
help("print")
is.character(cars$speed)
vec <- cars$speed
is.character(vec)
vec
for(pasta in vec){print(pasta)} #paste does not work in a loop, have to use print()
for(pasta in vec){paste(pasta)}
for(pasta in vec){cat(pasta, "\n")} #now it will work - \n means end of line
```

```{r}
help("substring") #extract or replace substrings in a character vector
fn <- "tofu"
nchar(fn)         #since tofu is a 4-letter word, nchar=4
substr(fn, 1, 2)  ##extract part of the word
substr("cooked tofu", start=1, stop=2) #extract part of the word
mystring <- "dog cat duck"
substring(mystring, c(1,5,9), c(3,7,12)) #divide into separate words
```

+ Search for words/strings in a data set (could be an article f.ex., searching for key words)

```{r}
help(strsplit) #a space within "" in R does not signify that it is two words, the space will just be counted as one of the characters in the object
strsplit(x="cooked tofu", split= " ") #to split the "" object into two words based on space in between, break this string based on the defined separator
paste("one",2,"three", 4)
paste("one",2,"three", 4, sep=";")
help(which)
which(LETTERS == "R") #to see which element in a vector is true to be given value
which(ll <- c(TRUE, FALSE, TRUE, NA, FALSE, FALSE, TRUE)) #which means give me/return which elements in the vector are true
```

## Stringr
  + Another tidyverse package
  + Use for cleaning variables
  + 7 main verbs:
  
      + `str_detect` search for a specific text in your string
      + `str_count`
      + `str_subset`
      + `str_locate`
      + `str_extract takes part of word you're interested in`
      + `str_match makes sure this exists in the string you're looking for`
      + `str_replace`
  
  + _ is metagrammar in tidyverse package for naming functions
  
```{r eval=FALSE, echo=TRUE} 
library(stringr)
x <- c("why", "video", "cross", "extra", "deal", "authority")
str_length(x)
nchar(x[1])
str_c(x, collapse = ", ")
str_sub(x,1,2)
str_subset(x, "[aeiou]") #when you search for strings in r you define a regular expression. an app in R studio can help you figure it out.
#install.packages("devtools")
devtools::install_github("gadenbuie/regexplain") # search for patterns, pattern could be an integer
regexplain:::regexplain_addin() # add-in console pops up
```

* regexplain can be used to search for numbers, expressions, words, ++ in a sample of text or data

# Working with factor variables

* Create ordinal(ordered) factors
* Factor variables are categorical variables, either numeric or character variables
  + they are treated speically in statistical modelling
  + should be stored as factors
     + use `factor()` to convert categorical variables to factors
     + R stores factors as a vector or integer values with a corresponding set of          character values to use for each value when the factor is displayed
     
```{r}
data <- c(1,2,2,3,1,2,3,3,1,2,3,3,1)
var <- factor(data)
var
levels(var)    #levels are always characters
levels(var) <- c("low", "mid", "high") #levels can be indexed and labeled
var <- factor(data)
factor(data, labels=c("I", "II", "III")) #define labels using the labels argument, which belongs to the factor() function
nlevels(var)   #how many levels are there?
class(var)
mode(var)
```

* You can convert factors to numeric or characters again in order to perform arithmetic operations, using levels

```{r}
levels(b)[b]
as.numeric(levels(b)[b])  #this converts the factor to a character variable, and then to a numeric variable which looks simpler
```

## Ordinal factors

* Methods and functions treat ordered factors and factors differently
  + use the argument `ordered=TRUE`in the `factor()` function to create ordered factors, or use the `ordered()` function
  + an ordered factor has class c("ordered", "factor")
  + you must first turn ordered factors back to numeric factors before doing calculations
  
```{r eval=FALSE, echo=TRUE} 
ordered(var)
class(ordered(4:1))
(b <- factor(a, ordered=TRUE))
(b <- ordered(a))
```

## Manipulating factors

* If subsets of a factor are extracted, the subsets will retain all of the original levels, even if they do not exist inthe subset
  + should remove them
  
```{r}
lets <- sample(letters,size=100,replace=TRUE)
lets <- factor(lets)
table(lets[1:5])
table(lets[1:5,drop=TRUE]) #drop=TRUE argument removes unused levels
table(factor(lets[1:5]))   #passing the factors to the table() function does same
```

  + use the `exclude=` argument in the `factor` function to exclude a particular level
  + use the `cut` function to create factors from continuous variables
  + when combining variables which are factors, since the c function will interpret the factors as integers, they should first be converted back to their original values (using levels function), then catenated and converted to a new factor
  
```{r}
fact1 <- factor(sample(letters,size=10,replace=TRUE)) 
fact2 <- factor(sample(letters,size=10,replace=TRUE))
fact1
fact2
fact12 <- factor(c(levels(fact1)[fact1],levels(fact2)[fact2]))
fact12
```

# Closing

*That's it for now!*