---
title: "Day 1"
author: "J.M. Lengle"
date: "2022-09-28"
output: html_document
---

# **Introduction to statistics in R: 1**

# R as a language

* R is an *object-oriented* programming language
  + object-oriented: has classes (e.g. dataset, string, matrix, list, +)
  + an important step to being comfortable working in R is knowing *what* you're working with (i.e. what class)
 
* Like any language, R has grammar that must be followed in order to create meaningful 'sentences'
  + a string is a collection of text. in order to create a string you use `""`.
  + a command prompt is a 'verb' for an action you want executed. these verbs are followed by `()`.
  
* There is a large online community of R users
  + Q&A site: Stack Overflow
  + Find and share codes: GitHub
  
Let's get started:

```{r eval=FALSE, echo=TRUE}    
print("hello World")          #ask R to print a string
setwd("~/Desktop/my r")       #set your working directory
getwd                         #shows your current working directory
source("helloworld.R")        #open a file in R
install.packages("tidyverse") #install the tidyverse package
library("tidyverse")          #load the tidyverse library
help("help")                  #type help(__) to look up help file
```

# Working in R

* Workspace panes
  + the top left pane is where you write your code
  + the bottom left pane, the R console, is where you run your code
  + the top right pane shows your code history and your "environment"
    + the environment includes your active data and values/variables/objects
    + your global environment = the active data and objects that are currently available to you during your session

| R panes 	|  	|
|---	|---	|
| code 	| important stuff (data files, variables) 	|
| results 	| plots, help files, r markdown output 	|

```{r}
library(pander)             #pander is helpful for generating tables
data(cars)
pander(cars)
```

## Creating data frames
* Data frames in R have 2 dimensions: rows and columns

```{r}
help("data.frame")
name <- c("Emily", "Julie", "Stine", "Eirik", "Nadine", "Sara", "Ole", "Anders", "Niklas", "Fredrik", "Haghish")
age <- c(27, 25, 31, 26, 31, 22, 27, 37, 44, 45, NA)
gender <- c("female", "female", "female", "male", "female", "female", "male", "male", "male", "male", "male")
df <- data.frame(name, age, gender)
pander(df)
```

## Running calculations
  + R can be used as a calculator
  + be clear on what you want by using parentheses cleverly
  + to create variables, use the `assign()` function
```{r}
2+2
2*5
1+2*5+1/2
help("assign")
assign("A", 10)                    #assign A=10
A                                  #request value of A
2+A
assign("B", 11)
B
```
 
  + variables can be single numbers, strings, vectors +
    + to make a vector, use the `c()`function
      + type of data: number, character/string

```{r eval=FALSE, echo=TRUE} 
x <- c(1:10)                      #assign vector 1-10 to x
x
TF <- c(TRUE, TRUE, T, FALSE, FALSE, F)
TF
```
  
  + operators in R
  
| Operators 	|
|---	|
| + 	|
| - 	|
| * 	|
| / 	|
| ^ or ** 	|

   + R also makes use of logical operators that produce true (1) or false (0) statements
   

| Logical operators 	|
|---	|
| < 	|
| <= 	|
| > 	|
| >= 	|
| == 	|
| != 	|
| !x 	|
| x | y 	|
| x & y 	|

### Subsetting data

* There are multiple ways to subset data

```{r eval=FALSE, echo=TRUE} 
data(iris)                         #specify data you want to use
print(iris[10:12,])                #print only observations 10-12
pander(iris, n=4)                  #print only observation number 4
View(iris)                         #view the data
View(iris[10:15,])                 #view only observations 10-15    
pander(iris[12:14,2:5])            #create a table of observations 12-14,2-5
iris[c(10,13,33,1,1),c(5,2,3,2)]   #use brackets to create an index of which rows and colums you want to view [index 1(row), index 2(column)]
View(iris)
rm(iris)                           #stop working with a data set, or remove  a variable from global environment
index <- iris$Sepal.Length < 5     #create a variable that represents a train of T/F statements for each observation in the data set concerning whether the value is <5 or not
                                   #use the $ to specify a column in a data set
length(index)                      #request length of variable
data(cars)
help("head")                       #head function extracts first values of data set
head(cars)
head(cars, n=6)                    #specify extract 6 first values
head(cars, n=6L)                   #L specifies extract integers
tail(cars)                         #extract last values of data set
x[(x>8 | x<5)]                     #ask R to list the values of x that are above 8 OR below 5
dim(cars)                          #retreive or set the dimension of an object
```

#### Let's do some calculating! woo

```{r}
print(df[df$gender == "female","age"])      
mean(df[df$gender == "female","age"])
                                   #what is the mean age of females in the sample
femmean <- mean(df[df$gender == "female","age"]) 
                                   #assign the mean to an R object
sd(df[df$gender == "male","age"])  #what is the standard deviation of the age of males in the sample
sd(df[df$gender == "male","age"], na.rm = TRUE)
malemean <- sd(df[df$gender == "male","age"], na.rm = TRUE) 
                                   #assign the sd of the mean of male age to variable malemean, and specify removal of missing values (na.rm=TRUE)
val <- df$age >= 35                #add a T/F variable that assesses whether the observed age of the participant is greater than or equal to 35
df$age35 <- val
pander (df)
hist(cars$speed)                   #generate a histogram of the speed of the cars
```

# R markdown

* R markdown is a tool for
   + writing code
   + documenting code
   + commenting/explaining code
 
* R markdown
   + is a form of literate programming: uses human language inside computer programming
   + accesses and uses the local environment (v. global), and thus provides a clean slate for analyses and actions in R
   + allows for the transformation of raw data --> reproducible data
   + can be used for mathematical notation
 
* Some benefits of R markdown are:
  1. Tests in new environment
  2. Easy to share outside R
  3. Produces automatic reports
  4. Can combine multiple programming languages
  5. Includes a table of contents
  6. Can run live analyses
  
* To use R markdown:
    + type human language in open spaces
    + use 'chunks' to insert R code
       + create a code chunk with `ctrl+option+i`
    + press knit to compile document to html, pdf or word document

## Formatting in R markdown

### Tips and tricks

* See R markdown cheatsheet and reference sheet for tips on how to format R markdown documents
* Some notes:
  + use #https://www.tablesgenerator.com/markdown_tables# to generate tables
  + change chunk opening to {r, eval=FALSE, echo=TRUE} if you would like the markdown document to show the code but not the results
        + setting eval=TRUE will provide results, and echo=FALSE will remove original code
  + exclude a part of R markdown using < ! -- >
  
# Closing

We learned a lot more than this summary, but you know that! *Takk for nå.*
