---
title: "Day 1 - Session 2"
output:
  html_document: default
  word_document: default
date: '2022-09-26'
---
Intro to RMarkdown 
=======================================================================================================================
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You can create headings by having either multiple = or - under the line you want as heading, or using one or multiple hashes. Amount of hashes defines the size of the header

# Heading 1

Heading 2 
--------------------------------------------------------------------------------------------------------------------------

### Heading 3

#### Heading 4


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.



# Monday Morning

We learned about vectors and so. 
You can create "chunks" to write R code by pressing cmd +  option + i. You write the code between the accents
```{r message=FALSE}
print("hello World")
library(tidyverse) #Load tidyverse package

assign(x = "A", value =10)
A
print(10, 20, 30)
print(c(10, 20, 30, "psy9510"))
B <- c(1, 2, 3, 4)
S <- c("hi", "hello")
X <- (1:10)

x <- c(1:10)
x
x[(x>8) | (x<5)]

#True or false are boolean operator which can either be true or false
TF <-c(TRUE, TRUE, T, FALSE, FALSE, F)
TF
typeof((TF)) #Type of gives us the data type of the variable
length(TF) #
```
Creating a list
---------------------------------------------------------------------------------

- List 1
- List 2
  - List 3
  - List 4

If you want to create a table you can use https://www.tablesgenerator.com/markdown_tables to generate a table to copy into RMarkdown. 

Or you can use the package Pander 

| ID 	| Hat 	| Glasses 	|
|----	|-----	|---------	|
| 1  	| 1   	| 0       	|
| 2  	| 0   	| 0       	|
| 3  	| 1   	| 1       	|


Working with dynamic R tables
========================================================================================================================


As you see, Rmarkdown's tables are stupid and we want a smarter way for working with dynamic tables. For this we use `pander` R package. 

```{r}
library(pander)

data(cars)
pander(head(cars))
pander(tail(cars, n = 3L)) #L makes it an integer, we can simply remove it. It does not change anything

```


```{r}
data(iris)
pander(iris, n = 4)

```


Creating a dataset
=========================================================================================================================

We create a data frame using the `data.frame`function

```{r}
help("data.frame")

```

Let's print all the english letteres. it's defined in R automatically and we do not need to define it ourselves

```{r}

LETTERS[1:3]
```

Subsetting data in R
=====================================================================================================================================

```{r}
data(iris)
print(iris[10:12,])

```

select the rows 12:14 and columns 2:5


```{r}

pander(iris[12:14, 2:5])
```
You can also be very selective in indexing 

```{r}
pander(iris[c(10, 13, 33, 1, 1), c(5, 2, 3, 2, 5, 5, 5, 1, 3)])

```

Creating a date.frame
==================================================================================================================

```{r}
name <- c("Emily", "Julie", "Stine", "Eirik", "Nadine", "Sara", "Ole", "Anders", "Niklas", "Fredrik", "Hagish")
age <- c(27, 25, 31, 26, 31, 22, 27, 37, 44, 45, NA)
gender <- c("female", "female","female", "male", "female", "female", "male", "male", "male", "male", "male")
df <- data.frame(name, age, gender)
pander(df)
```

Question: What is the mean of age for female participants of the class PSY9510

```{r}
print(df[df$gender == "female","age"])

mean(df[df$gender == "female","age"])

```

Lets assign the mean to an R object

```{r}
femmean <- mean(df[df$gender == "female","age"])
```

The analysis showed that the mean age of females in this dataset is 
`r femmean`. **This way we can bring R objects inside of text**

Let's calculate the standard deviation for male's age:

```{r}
#help(sd)
sd(df[df$gender == "male", "age"], na.rm = TRUE)

```


How can I evaluate if the age of the participants is abaove or equal to 35 or not

```{r}
val <- df$age >= 35

#now I define a new variable
df$age35 <- val

pander(df)
```



We can add figures to R markdown. Here I will add a historgram using the `hist`function. 
```{r}
hist(cars$speed)

```



