---
title: "Day2"
author: "L. Mayerhofer"
date: '2022-09-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Getting started

Installing and loading packages. To do that, we use `install.packages("packagename")` followed by `library(packagename)`

```{r}
library(readstata13)
library(foreign)
library(haven)
library(readxl)
```

data(airquality)

##Loading df with different extensions:
For that, we use `read_fileextension("name of the file")`

##Class of an object
Use comands `is.factor` or `class()`

#Cleaning environment

Use `rm(list=ls())`

#Subseting
data -column
Colnames()


#Saving files

Use `saveRDS(datasetname, "datasetsaved.RDS")`

```{r}

data(iris)
saveRDS(iris, "testdata.RDS")
```

##Save as csv

Use `write.csv` 

```{r}
data(iris)  
write.csv(iris, file="iristest.csv")

```


#Looping

Use the funtion `for`

```{r}
for(b in 1:5) {print(b)} 
s <- c("some", "character", "for", "the", "example")
for(indiancurry in s) {print(indiancurry)}
print(s[3])