---
title: "test"
author: "L. Mayerhofer"
date: '2022-09-26'
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# R Markdown

# Heading

To write headings at an R markdown, we add `#`. If we add additional `#`, we have subheadings.

## Example 1

### Example 2

#### Example 3


#Monday Morning

We learned about vectors and so!
```{r, message = FALSE, warning = F}
print("hello world")
library(tidyverse)
assign (x = "A", value =10)
A
print(c(10, 20, 30, "psy9510"))
```

##Creating a list


#Working with dynamic R tables
As you see, Rmarkdown's table are stupid and we want a smarter way working with dynamic tables. For this, we use the `pander` R package

```{r}
library(pander)
data(cars)
pander(cars)
pander(head(cars, n = 6L))
pander(tail(cars, n =3L)) # L just means an integer, we can simply remove it.
```


#Create a dataset
For that, we use `data.frame`
```{r}
help("data.frame")
```

## Printing English letters.
These are already defined in R. 
```{r}
LETTERS
L3 <- LETTERS[1:3]
```

#Subseting
```{r}
data(iris)
sub1 <- iris[10:15,]
sub1
```
```{r}
pander(iris[12:14, 2:5])
```

