---
title: "psy9510 Day 2"
author: "Nadine Farnes"
date: '2022-09-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This markdown document gives examples of different functions in R learned in the course PSY9510 on day 2.

# FOR-LOOP IN R
We can use loops for performing repetitive tasks. The loops repeat operations a specified number of times. We can use the `for loop` to loop a function over a vector or matrix. This saves time and processing power, and is an efficient way to perform a function without using many lines of code.  
First, we state an index _b_ which equals 1 in the first iteration and 5 in the last iteration.  
Then we state the function we want to be repeated, here `print()`. 
```{r}
for(b in 1:5) {       
  print(b)            
  }
```

Below is another example with a vector containing strings. Here we specify that _i = S_, which takes on the first element of _S_ in the first iteration, and continues through the vector until _S_ has reached its length. We print _i_ in each iteration, meaning we print each element in the vector.
```{r}
S <- c("some", "character", "for", "the", "example")

for(i in S) {
print(i)
}
```

Another way to achieve the same output is given below. Here _b_ takes on values from 1 to 5. _print(S[b])_ means that we print the _bth_ character from _S_, again printing each element of the vector.
```{r}
for(b in 1:5) {
  print(S[b])
}
```

# Read and save datafiles

We can read datafiles from the working directory or save a dataframe to the current working directory using `readRDS()` and `saveRDS()` respectively.
```{r}
data(iris)                           # load the data
saveRDS(iris, "testdata.RDS")        # save the RDS file to the current working directory
df2 <- readRDS("testdata.RDS")       # read the RDS file and assign to a new variable called df2
```
 
 
## CSV format
CSV formats are dataformats usaully separated by a comma or semicolon, and where every row of the data ends with a new line.  
To read the CSV format files we can use the funtion `read.csv()`, and save to the current working directory we can use `write.csv()`.

```{r}
write.csv(iris, "iristest")         # write to .csv format
iris2 <- read.csv("iristest")       # read -csv format and assigning to a new variable iris2
```

## SPSS format

Reading and writing SPSS datafiles can be facilitated through the packages `foreign`and `haven`.
We can write an SPSS file from an R object using `write_sav()`, and read an SPSS file into R using `read_sav()`.  
We can use `::` to access the specific function from the haven library.

```{r}
library(foreign)                                            # load library foreign
library(haven)                                              # load library haven
data("airquality")                                          # load dataset
haven::write_sav(data=airquality, path="airquality.sav")    # the data airquality is written to a .sav file with the filename given by path
df <- haven::read_sav(file = "airquality.sav")              # reading the .sav file and assigning it to a new variable
```

> Remember to be careful with variables that have labels. They might be converted to factor (nominal) variables in R, even if they are stored as numeric in SPSS. We can specify this in `read.spss()` using argument _use.value.labels = FALSE_

## STATA format

We can write data to stata binary format and read using the libraries `foreign` and`readstata13`. 
`write.dta()`writes to a Stata format while `read.dta13()`reads a Stata .dta file and imports the data into a dataframe.  
We can specify if we do not want R to make factors out of the labels by using the argument _convert.factors = FALSE_.
<!-- continue HERE!!!! -->

```{r}
library(readstata13)                                         # load library readstata13
foreign::write.dta(airquality,"airquality.dta")              # the data airquality is written to a .dta file
df2 <- readstata13::read.dta13("airquality.dta")             # reading the .dta file and assigning it to a new variable
```


## EXCEL format

We can use the `read_excel` package to read excel files to R.

find the examples in the library

```{r}
library(readxl)                                               # load the library readxl
readxl_example()                                              # showing the example files that read_excel has
xlsx_example <-(readxl_example("datasets.xlsx"))              # assigning the dataset to a new variable 
xlsx_example2 <- read_excel(xlsx_example)                     # reading the dataset and assigning it a new variable
```

## Worksheets
We can specify which worksheets we want to access in the excel file. First we can list the sheet names using `excel_sheets()` and then we can specify which sheet we want to work with using the argument _sheet_ in `read_excel()`.
```{r}
excel_sheets(xlsx_example)                                    # list sheets
read_excel(xlsx_example, sheet = "chickwts")                  # specify sheet by name
read_excel(xlsx_example, sheet = 4)                           # specify sheet by number
```

## Reading cells

<!-- KONTROLLER DENNE KODEN! -->


We can also control which cells are read in the excel file by using arguments of `readexcel()`.
```{r}
read_excel(xlsx_example, n_max = 3)                      # Read a maximum of 3 rows
read_excel(xlsx_example, range = "C1:E4")                # Read cells C1 o E4
read_excel(xlsx_example, range = cell_rows(1:4))         # Read only cell rows 1 to 4
read_excel(xlsx_example, range = cell_cols("B:D"))       # Read only cell columns B to D
read_excel(xlsx_example, range = "mtcars!B1:D5")         # Read only from the sheet called mtcars and cells B1 to B5
```

Furthermore, we can specify NAs if they are represented by something other than blank cells. Here, for example we set all cells named _setosa_ to NA.  
```{r}
read_excel(xlsx_example, na = "setosa")
```


# Cleaning the workspace/working directory
 
We can use `ls()` to list the variables in your working directory/global environment. If you have an empty environment, it returns 0. Then we can use `rm()` to clear the workspace.  
```{r}
rm(list = ls())
```

We can also remove files from the working directory using `unlink()`.  
```{r}
unlink("iristest.csv")
```
 

# Subsetting

We have previously learned that we can use brackets _[]_ to access parts of a dataset. We can also subset vectors, matrices or dataframes using the `subset` function.
```{r}
head(subset(airquality, Temp > 80, select= c(Ozone, Temp)))  # Subset (showing head of dataset) where temperature is above 80 and only with columns Ozone and Temp
subset(airquality, Day == 1, select = -Temp)           # All observations of day 1, but excluding the temperature variable
subset(airquality, Day == 1, select = -c(Temp,Wind))   # Same, but excluding temperature and wind variables
head(subset(airquality, select = Ozone:Wind))          # Head of observations of columns Ozone to Wind
```




# Names of rows and columns
`colnames`gives column names of the matrix, while `rownnames`gives rownames of the matrix
```{r}
colnames(airquality)
rownames(airquality)    #airquality does not have rownames so the function lists the rownumbers
```


# Checking class of data
Previously, we learned that we can use the functions `str()`and class `class()` to check what datatype a vector contains. We can also check for each data type specifically using `is.factor()`, `is.character()` and `is.numeric()`, returning _TRUE_ or _FALSE_.

```{r}
is.factor(iris$Species)                        # check if a variable is a factor
is.character(airquality$Temp)                  # check if a variable is a character
is.numeric(airquality$Temp)                    # check if a variable is a numerical
```


# Summary 
`summary()`gives a summary of the different variables in a dataframe, such as descriptive statistics of mean, median, minimum and maximum values. It also shows number of missing (NAs). We can also check missing with `is.na()`and `sum()`.
```{r}
summary(attenu)                                # summary of dataframe
sum(is.na(attenu$station))                     # sum of the counts of missing in the variable
```

# If else statments
We can make conditional statements in R using `if(){}`and `else(){}`. We read it as: "If X is TRUE do do this, if false do something else".  
The general form of the code is shown in pseudocode below:
```
 if (condition) {
  Code executed when condition is TRUE
  } else {
  Code executed when condition is FALSE
  }
```

[Source](https://intro2r.com/conditional-statements.html)

In the below example we say: if the variable event is numeric, calculate the mean of this variable. If not, print the structure of the variable. Since the first conditional statement is _TRUE_, the mean is printed.
```{r}
if(is.numeric(attenu$event)){
  mean(attenu$event)
} else {
  str(attenu$event)
}
```

# Working with strings

## Subset of strings
We can subset strings using `substr()` and `strsplit()`. 

```{r}
fn <- "cooked tofu"                                   # defining a new variable fn
nchar(fn)                                             # find number of characters of fn
substr(fn,8,10)                                       # give the 8th to 10th characters of the string
substring(state.name, 1,3)                            # give the first three letters of each state

strsplit(fn, " ")                                     # splits the string into two, the split being defined by the space 
strsplit(fn, "")                                      # splits the string for all characters in the string
            
```

## Combining strings

We can use different printing functions to combine character strings: `paste()`, `cat()`, `paste0()`

```{r}
vec <- colnames(attenu)                          # a new variable containing column names of attenu
paste(vec[1], vec[2], length(vec))                # concatenating string and numeric value into one string using paste()
paste0(vec[1], vec[2], length(vec))               # same as above but with no spaces between

for(pasta in vec) {                               # printing the elements of vec using a for-loop and cat() (with spacing)
  cat(pasta, " ")
}

for(pasta in vec) {                               # same as above but with a line break between
  cat(pasta, "\n")
} 
```

## Which
`which()` finds the position where the value equals TRUE

```{r}
LETTERS == "R"                                               # gives a logical array, where TRUE if equals R
which(LETTERS == "R")                                        # finds indices of the where the Rs are, here 18 
ll <- c(TRUE, FALSE, TRUE, NA, FALSE, FALSE, TRUE)           # make new logical array
which(ll)                                                    # find indices which equals TRUE
```

## Searching for terms in string
`stringr` package has several functions to search for specific terms in string
`str_count()`, `str_subset()`, `str_extract()`, `str_replace()`

```{r}
library(stringr)
x <- c("why", "video", "cross", "extra", "deal", "authority")   # new variable with string elements
nchar(x[1])                                                     # how many characters does the first item have?
str_length(x)                                                   # how many characters do all the elements have?
str_count(x)                                                    # same as above
str_subset(x, "[a]")                                            # finds all elements with "a" in them
str_extract(x, "a")                                             # returns NA if "a" is not found, "a" if found
str_replace(x, "a", "e")                                        # replaces all a's with e's

tex <- ("I really have to leave in 2 min or 3")                 # a string with numbers in them 
as.numeric(unlist(str_match_all(tex, "[0-9]+")))                # finds all numbers in strings. Unlist() makes them into separate unlisted elements, as.numeric turns them into numerical values

```
