---
title: "R-kurs day 2"
author: "Niklas Karlsen"
date: "2022-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

### Conversion: 
as.? (`as.character()`, `as.factor()`, ...)  

### Boolean: 
is.? (`is.character()`, `is.numeric()`, ...)

```{r}
which(LETTERS == "R")                       # which position does R have?
```


### Loops: 
looping over columns or rows, and modifying values.

```{r}
vec <- c(4, 7, 12, 13)
for (v in vec) {print(is.character(v))}
```


### Different print-functions: 
`paste()` (converts to character and concatenates), `paste0()`, `cat()` (\n for newline)

### String-manipulations: 
`substring()` (extract or replace substring), `strsplit()` (split string)

```{r}
fn <- "tofu"
nchar(fn)
substr("cooked tofu", start=8, stop=12)
strsplit("cooked tofu", " ")
```

```{r}
library(stringr)
x <- c("hei", "cheese", "tofu")
str_length(x)                                     # length of each word in x
# str_c(x)
y <- str_subset(x, "e")                                # selects words that contains "e", uses regular expressions
```

The following words contain the letter e: 

```{r} 
print(y)
```

## References for packages used

```{r} 
citation("stringr")
```
