---
title: "Day 1 - R-course - Sara"
author: "Sara Berger Marjanovic"
date: "2022-09-28"
output:
  word_document: default
  pdf_document: default
  html_document: default
---
# R and RStudio

R is an object-oriented language

The code is typed in the consule
The global environment is where your data, functions and objects are stored.

A function can be compared to a verb - it is what is being done to the object inside the parantecies

# Seeking help

Knowing how to seek help is a big part of programming! No one have ALL the answers!
Action plan:

1. Google it!

2. Use ´help()´-function in R to figure out what arguments a function take. 

3. Describe you problem as clear as possible an put it on to StackOverflow

# Preperations

## Specifying working directory

When we set a working directory we tell R in what folder we are working in. 
We can set our working directory by using the function:

```{r}
#setwd(/Users/saramarjanovic/OneDrive - Universitetet i Oslo)
```

I have know set my working directory to my One Drive, in no specific folder. 

## Installing and loading packages

In R we use the function:

```{r}
#install.packages("nameofpackage")
```
in order to install open source packages in R.
 
A very useful and well known collection of packages is the Tidyverse.

In order to load the package we use the function:

```{r}
library(tidyverse)
```

Here I have loaded the Tidyverse-package and I can now use Tidyverse-functions!

# The ´help-function´

The help-function is a tool to read the help files of different packages or functions in R

We can even use the help-function to learn about the help-function:

```{r}
help(help)
```

Description from help file on ´help()´
"Help is the primary interface to the help systems."
The help files can be confusing, but are helpful sometimes. 

# R as a calculator

We can use R as a calculator:

```{r}
2+2
2-1
2*3
4/3
```

R follows the order of arithmetic operations, which is important to remember if we want to perform more complex mathematic operations. 

# Assigning values

```{r}
help(assign)
```

Description:
Assign a value to a name in an environment.

We can assign values so that they show up on our global environment in different ways:

By using the `assign`function

```{r}
assign("a", 20)
```

or by using the `= ` sign or the `<--`sign. 

Like this:

```{r}
b = 10
d <- 15
```

# Vectors

A vector is a one dimensional object with a train of elements. A vector is an anatomical object, meaning every element in that object needs to be of the same type!
We specify to R that our object is a vector by using the keyword ´c´ and seperate each element with a comma, like this:

```{r}
my_vector=c(10,20,30)
class(my_vector)
typeof(my_vector)
```

my_vector is a numerical vector, because all the elements are numbers.

If we include a string as an element in the vector, the vector will become a character vector. 

```{r}
my_chr_vector = c(10,20,30, "R is diffcult")
typeof(my_chr_vector)
class(my_chr_vector)
```

We can also assign vectors with the assign-function like this:

```{r}
assign("another_vector", c(10,20,30) )
```

The ´:´-siggn tells R that we want every value from x:y. We can use this when assigning vectors, instead of writing every number from x to y, like this:

```{r}
my_easier_vector = c(1:10)
```

In this case we could also write it without the c-keyword, the output will be the same.

```{r}
my_even_easier_vector = 1:10
```

# Indexing and subsetting

We can subset by using implying the position (index) we would like to subset from inside the square bracket, in this instance after our vector, like this:

```{r}
ex_vector = c(3, 1, 4, 1, 5, 9, 2, 6, 5)
ex_vector[4]
```

We can also subset different intervals by using and (&)/or(|) statements, like this:

```{r}
ex_vector[(ex_vector>8) | (ex_vector<5)] 
ex_vector[2&6]
```

From the first example we will expect the output of the elements that are over 8 and under 5.

For the second example we will expect the output of the element on the second position and the sixth position. 

NB! R counts form 1, unlike many other languages that counts from 0.

# Boolean data types and logical operators

## Boolean variables

Boolean variables only take TRUE or FALSE statements

```{r}
TF = c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE)
typeof(TF)
class(TF)
length(TF)
```

       
## Logical operators

Logical operators return TRUE or FALSE:

Greater than (>)
Less than (<)
Greater than or equal to (>=)
Less than or equal to (<=)
Equal to (==)
Not equal to (!=)

Example:

```{r}
my_variable = 10
my_variable > 20  #will give FALSE because 10 is less than 20.

my_variable != 20 #will give TRUE because 10 is not equal to 20

```

# Data frames

You can create a data frame by using the data.frame-function, importing a file, or by using one of the data frames already avilable in R. 

## Creating a data frame

Data frames are two-dimensional objects. This means we can combine two one-dimensional objects (eg. vectors) into a data frame, like this:

```{r}
#vector 1
TV_shows = c("Game of Thrones", "Luther", "Orange is the New Black" )

#vector 2
good = c(TRUE, TRUE, FALSE) #pls don't fail me if you like Orange is the New Black 

#vector 3
fantasy = c("yes", "no", "no")


my_personal_opinion = data.frame(TV_shows, good, fantasy)
```

We can index data frames by using the dollar sign, like this:

```{r}
my_personal_opinion[my_personal_opinion$good==TRUE] #what did I do wrong here? It's only supposed to show the good once :/
```

# Creating tables

We can create tables directly in R, but they are kinda ugly. With the ´pander´-package we can create more beautiful tables in RMarkdown!

```{r}
#install.packages("pander")
library(pander)
help("pander")
```

Let's say I want to make a table with the iris data frame, showing petal length, petal width and species:
```{r}
data(iris)
pander(table(iris$Petal.Length, 
              iris$Species ))

```

We can also use Markdown Tables Generator to make it even easier:
[https://www.tablesgenerator.com/markdown_tables] 

# Creating plots

We can create plots in R by using plot-functions.
The plot-function gives a scatterplot, but you can also specify other blots like barplots, boxplots and histograms. 

Let's make a scatterplot from the iris data frame:
```{r}
plot(x=iris$Petal.Length, y=iris$Petal.Width)
```
It doesn't make a lot of sense, but it is what it is!



