---
title: "R-homework Tuesday Sara"
author: "Sara Berger Marjanovic"
date: "2022-09-26"
output:
  pdf_document: default
  html_document: default
---


# Loops

## What are loops?

R-loops can be used to execute statements (eg. mean, summary, paste) over a number of element in an R-object. We can also set limitations to which elements should be included. 

## Creating loops

### For-loops

A for loop in R can be used to iterate over the elements of an R object

Example:

```{r}
for(b in 1:5) {
  print(b)
}
  
q = c('some', 'character', 'for', 'the', 'example')
for(indiancurry in q) {
  print(indiancurry)
}


```

### If-loops

If/else-loops are also known as conditionals, because we can apply conditions to our loop. 
If this condition is met -> we do this
If else -> we do this instead

```{r}
#I don't know what went wrong here :/
#my_mean = function(my_vector){
  #if(class(my_vector)=="character")
   #print("This is not a numeric vector, dummy!")
#} else {
  #sum_vector=0
  #for(element in my_vector)
    #sum_vector = sum_vector + element
#}

#mean_vector = sum_vector/length(my_vector)
#return(mean_vector)
```

# Working with data

## Loading data into R

When you load a file into R, you need to specify the file path

We have different open source packages for loading different data files into R.
For Stata, xlxs., SPSS, and CSV data there are different packages and functions.

### CSV data format

```{r}
help("read.csv")
```

Description:

Reads a file in table format and creates a data frame from it, with cases corresponding to lines and variables to fields in the file.
File can also be a complete URL!

```{r}
help("write.csv")
```

Description:

Write.table prints its required argument x (after converting it to a data frame if it is not one nor a matrix) to a file or connection.

```{r}
write.csv(iris, file="irisdata.csv")
```

### Reading Stata data

#### Read.dta13

read.dta13 reads a Stata dta-file and imports the data into a data.frame. readdta13 is part of the readstata-package. 

```{r}
library(readstata13)
help("read.dta13")
library(haven)
library(foreign)
```

:: tells R we are using the command from this exact library

### Reading SPSS files

#### Read SPSS

```{r}
#read.spss()
help("read.spss")
```

read.spss reads a file stored by the SPSS save or export commands.

reas.spss was orignally written in 2000 and has limited support for changes in SPSS formats since (which have not been many - LOL).

### Reading xlsl. files

```{r}
library(readxl)
help("readxl")
```

read_excel() calls excel_format() to determine if path is xls or xlsx, based on the file extension and the file itself, in that order. Use read_xls() and read_xlsx() directly if you know better and want to prevent such guessing.
Sheet to read. Either a string (the name of a sheet), or an integer (the position of the sheet). Ignored if the sheet is specified via range. If neither argument specifies the sheet, defaults to the first sheet.
xl-sheets are complicated. This package enables you to specify sheets and columns

Example:

```{r}
#read_excel(xlsx_example, sheet = "budget_this_year")
#read_excel(xls_example, sheet = 4)
```


### RDS-functions

We can also use the rds-functions to save an R-object to a file, and then restore it. 

.rds is a binary 


```{r}
# saveRDS()
# readRDS()
```

#### readRDS

```{r}
#readRDS(object, file = "file name")
```


#### saveRDS

```{r}
# saveRDS(iris, "testdata_Tuesday.RDS")

```

saveRDS only send one file

## Cleaning workspace

### Removing all objects in global environment

```{r}
#rm(list=ls())

s=34
rm(s)
```

### Some handy functions to get an overview of your dataframe

dim() gives the dimensions of your R-object (eg. a data frame)

str() gives an overview of the the internal structure of a R object. If we use the str-command on a data frame we will get the class of every column, and this can be handy if we want to know what functions we can use on the different columns. 

subset() is another way to subset a dataframe. We can select the rows and columns that we would like to look at. But we can also subset by using square brackets with positions or the dollar-sign.

```{r}
data("airquality")

dim(airquality)
str(airquality)
subset(airquality, Day==1, select = -Temp)
```

### Column- and rownames

Retrieve or set the row or column names of a matrix-like object, like this:

```{r}
#rownames()
#colnames()
```

If we would like to set new columnnames we can create a vector with new columnnames and set them as such.


## ´is.na´-function

```{r}
help("is.na")
help("as.na")
```
NA is a logical constant of length 1 which contains a missing value indicator. NA can be coerced to any other vector type except raw. There are also constants NA_integer_, NA_real_, NA_complex_ and NA_character_ of the other atomic vector types which support missing values: all of these are reserved words in the R language.

The generic function is.na indicates which elements are missing
The generic function is.na<- sets elements to NA.
The generic function anyNA implements any(is.na(x)) in a possibly faster way (especially for atomic vectors).


# Character vectors

Character vectors are vectors with character-elements in them. 
Example:

```{r}
cool_chr_vector = c("You", "are", "very", "cool", "and", "very", "loved")
```

## Paste-function

```{r}
help("paste")
```

Paste converts its arguments (via as.character) to character strings, and concatenates them (separating them by the string given by sep). If the arguments are vectors, they are concatenated term-by-term to give a character vector result. Vector arguments are recycled as needed, with zero-length arguments being recycled to "" only if recycle 0 is not true or collapse is not NULL.

Example:

```{r}
paste("Hello", "I", "am", "the", "judge", "Ola", "is", "number", 1)
```


## as.character-function

```{r}
help(as.character)
```
as.character and is.character are generic functions that reate or test for objects of type "character".

Example:

```{r}
wrong_vector = c(1,2,3,4,5)
as.character(wrong_vector)
is.character(wrong_vector) #now true because I made it into a character-vector
```

We can write methods to handle specific classes of objects

```{r}
q = c(1:50)

for (pasta in q) {
  print(pasta)
}
```

## cat-function

The cat-function takes different elements, convert them into character-types, combine them and prints them.

```{r}
help(cat)
cat("I", "am", "learning", "R")
```
Outputs the objects, concatenating the representations. cat performs much less conversion than print.

# Manipulating character data

```{r}
help(substring)
help("strsplit")
```

### Substring

Extract or replace substrings in a character vector.

Arguments
x, text	a character vector.

start, first	integer. The first element to be extracted or replaced.

stop, last	integer. The last element to be extracted or replaced.

value	a character vector, recycled if necessary.

```{r}
substr("cooked tofu", start=1, stop=2)
```


### Stringsplit

Split the elements of a character vector x into substrings according to the matches to substring split within them.

```{r}
strsplit(x="cooked tofu", split = " ")
```

```{r}
substring(state.name,1,3)
```

## which-function

```{r}
help(which)
```

Give the TRUE indices of a logical object, allowing for array indices.

```{r}
LETTERS == "R"
which(LETTERS == "R")
```

## Stringr-package

Stringr is a tidyverse-packeage that can help with data cleaning and string-manipulations.
Stringr-functions take character-vectors as first argument.

Strinr-cheetsheat: [https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf]

There are seven main verbs that work with patterns:

str_detect(x, pattern) tells you if there’s any match to the pattern.


str_count(x, pattern) counts the number of patterns.


str_subset(x, pattern) extracts the matching components.


str_locate(x, pattern) gives the position of the match.


str_extract(x, pattern) extracts the text of the match.


str_match(x, pattern) extracts parts of the match defined by parentheses.


str_replace(x, pattern, replacement) replaces the matches with new text.


str_split(x, pattern) splits up a string into multiple pieces.


## Regexplain-package
